### 2024-10-20 src/Submission.Rnw
Things 3

% Special Unicode characters
\newfontfamily\greekfont{DejaVu Sans}[Script=Greek]
\newunicodechar{α}{{\greekfont α}}
\newunicodechar{β}{{\greekfont β}}
\newunicodechar{γ}{{\greekfont γ}}
\newunicodechar{μ}{{\greekfont μ}}

@@

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

\section{Introduction}
The importance of understanding the diabetes dataset it to further our knowledge of what the key causes of diabetes are, and how they can be predicted. The dataset contains 11 variables, and the target variable is the progression of diabetes. The dataset contains 442 observations and 11 variables and we are understanding the factors involved which are:
\begin{itemize}
    \item \textbf age    - age in years
    \item \textbf sex    - male and female
    \item \textbf bmi    - body mass index
    \item \textbf bp     - average blood pressure
    \item \textbf s1     - TC: total serum cholesterol
    \item \textbf s2     - LDL: low-density lipoproteins
    \item \textbf s3     - HDL: high-density lipoproteins
    \item \textbf s4     - TCH: total cholesterol / HDL
    \item \textbf s5     - LTG: possibly log of serum triglycerides level
    \item \textbf s6     - GLU: blood sugar level
    \item \textbf target - measure of disease progression one year after baseline
\end{itemize}
Each of the feature variables have been mean centered and scaled by the standard deviation times the square root of the number of sample.

Our investigation is to look at the more commonly known variables that affect the progression of diabetes in the observations of this dataset. These are more specifically age and BMI. We will be using the maximum likelihood estimation method to estimate the parameters of the linear regression model for the progression of diabetes based on BMI, then using the model we will do a comparison study on how age combined with BMI affects the progression of diabetes compared to BMI alone.

We will compute 95\% confidence intervals for the parameters estimated in our MLE model using the likelihood interval obtained earlier. This will provide a range of plausible values for each parameter in the population, helping us understand the uncertainty in our estimates.

\section{Maximum Likelihood Estimation (MLE)}
<<echo=TRUE>>=
head(ddat)

@@

# Plot residuals to check for patterns
plot(ddat$bmi, residuals, main = "Residuals vs BMI", xlab = "BMI", ylab = "Residuals")
@
\subsection*{MLE Data Analysis}
With the assumption that the data is of a normal distribution and follows linear regression beta0 = 21.86, indicates that this is your progression if your BMI was 0, this is unrealistic as no one has this, the slope being labelled beta1 = 0.31 indicates every BMI increment of 1 would increase progression by 0.31 showing a small positive association which is reasonable, and standard deviation being 4102.63 is unrealistically high implying that BMI alone is not enough to determine the progression of diabetes based on the model or there is outlier variance that is just not explained by BMI. This is reasonable given the data is 10 measured variables and you would assume that the other 9 are not all redundant and cannot be predicted by BMI alone even though there is some correlation.

\section{Likelihood Ratio Test}
<<echo=TRUE>>=
# Fitting Model 1: Progression ~ BMI
model1 <- lm(target ~ bmi, data = ddat)

# Fitting Model 2: Progression ~ BMI + Age
model2 <- lm(target ~ bmi + age, data = ddat)

# Summarise the models
summary(model1)
summary(model2)

# Extract the log-likelihood for both models
logLik_model1 <- logLik(model1)
logLik_model2 <- logLik(model2)

# Print the log-likelihoods (optional)
logLik_model1
logLik_model2

# Calculate the likelihood ratio test statistic
LRT_stat <- 2 * (logLik_model2 - logLik_model1)

# Degrees of freedom is the difference in the number of parameters
df <- df.residual(model1) - df.residual(model2)

# Compute the p-value for the test
p_value <- pchisq(LRT_stat, df = df, lower.tail = FALSE)

# Print the result
cat("Likelihood Ratio Test Statistic:", LRT_stat, "\n")
cat("p-value:", p_value, "\n")
@
\subsection*{Likelihood Ratio Test Analysis}
Given the p-value is such a small number being less than 0.05, it can be concluded that the model with BMI and age is a better fit than the model with BMI as the only constituent and the likelihood ratio test backs up this data and indicates a large improvement. This was almost assumed given the data in the set has 10 variables and would be poorly constructed if so many were redundant which concludes through testing and general reasoning that adding additional variables will contribute to the accuracy of the model.

\section{Ranking the variables based on the correlation coefficient}
<<echo=TRUE>>=
# List of predictor variables (excluding the target variable)
predictor_vars <- c("age", "sex", "bmi", "bp", "s1", "s2", "s3", "s4", "s5", "s6")

# Calculate the correlation between each predictor and the target
correlations <- sapply(predictor_vars, function(var) {
  cor(ddat[[var]], ddat$target)
})

# Print the correlation coefficients
correlations

# Rank variables by absolute correlation
correlations_sorted <- sort(abs(correlations), decreasing = TRUE)

# Print the ranked correlations
correlations_sorted
@
This shows the most important variable is BMI, followed by possibly log of serum triglycerides level, and the blood pressure, which is a common factor in diabetes progression. The least important variable is gender as diabetes can manifest in any gender and makes it an irrelevant variable with minimal contribution to the dataset but more for record keep purposes.

@@

@@

>>>>

% Special Unicode characters
\newfontfamily\greekfont{DejaVu Sans}[Script=Greek]
\newunicodechar{α}{{\greekfont α}}
\newunicodechar{β}{{\greekfont β}}
\newunicodechar{γ}{{\greekfont γ}}
\newunicodechar{μ}{{\greekfont μ}}
\newunicodechar{θ}{{\greekfont θ}}
\newunicodechar{σ}{{\greekfont σ}}

@@

\begin{document}

@@

\section{Maximum Likelihood Estimation (MLE)}

We will use MLE to estimate the parameters of a statistical model that describes the relationship between BMI and diabetes progression. This will help us understand the strength and nature of BMI effect on the population.

<<echo=TRUE>>=
head(ddat)

@@

# Plot residuals to check for patterns
plot(ddat$bmi, residuals, main = "Residuals vs BMI", xlab = "BMI", ylab = "Residuals")
@

\section{Likelihood Ratio Test}

We will perform a likelihood ratio test to compare two nested models: one that includes only BMI as a predictor of diabetes progression, and another that includes both BMI and age. This will help us determine whether adding age significantly improves our understanding of diabetes progression.

@@

<<echo=TRUE>>=
# Log-likelihood function for the normal distribution
log_likely <- function(col, mu) return(- (1 / (2 * var(col))) * sum((col - mu)^2))

# First derivative of a of the log-likelihood function
score <- function(col, mu) return(sum((col - mu) / var(col)))

# Second derivative of a of the log-likelihood function
hessian <- function(col) return(-length(col) / var(col))

newton_raphson_mle <- function(col, θ = 1, tolerance = 1e-8, max_iter = 1000) {
    h <- hessian(col)
    for (i in 1:max_iter) {
        s <- score(col, θ)  # Call the score function
        theta_new <- θ - solve(h, s) # Update θ
        if (max(abs(theta_new - θ)) < tolerance) return(theta_new)
        θ <- theta_new
    }
    warning("Newton-Raphson algorithm did not converge")
    return(θ)
}

conf_int <- function(col, mle, conf_level = 0.95, range = c(-1, 1), tolerance = 1e-8) {

    h <- hessian(col)
    alpha <- 1 - conf_level

    # Normal-based confidence interval
    se <- sqrt(-1 / h)
    z <- qnorm(1 - alpha / 2)
    nm_min <- mle - z * se
    nm_max <- mle + z * se

    # Likelihood-based confidence interval
    max_ll <- log_likely(col, mle)
    crit_value <- qchisq(conf_level, df = 1) / 2
    root_find <- function(theta) log_likely(col, theta) - (max_ll - crit_value)
    ll_min <- uniroot(root_find, c(range[1], mle), tol = tolerance)$root
    ll_max <- uniroot(root_find, c(mle, range[2]), tol = tolerance)$root

    return(list(nm = c(min = nm_min, max = nm_max), ll = c(min = ll_min, max = ll_max)))

}

for (name in names(ddat)) {

    if (name == "target") next
    col <- ddat[[name]]
    mle <- newton_raphson_mle(col)
    ci <- conf_int(col, mle)

    cat(

        "\n\nColumn:", name, "\n",
        "Newton Raphson MLE: ", mle, "\n",
        "Normal CI: [", ci$nm[1], ", ", ci$nm[2], "]\n",
        "Likelihood CI: [", ci$ll[1], ", ", ci$ll[2], "]\n",
        sep = ""

    )
}
@

@@

<<echo=TRUE>>=
bootstrap <- function(data, lvls, α = 0.05, studentized = FALSE, debug = FALSE) {

    par_mean <- mean(data)
    lvls_len <- length(lvls)

    recurse <- function(data, lvls, α, par_mean) {

        len <- length(lvls)
        if (len == 0) return(list(μ = mean(data), std_err = sd(data)/sqrt(length(data))))

        n <- length(data)
        iter <- lvls[1]
        boot <- data.frame(
            μ = numeric(iter),
            se = numeric(iter),
            t_stat = numeric(iter)
        )

        for (i in 1:iter) {# Loop through each iteration at the current depth

            # Bootstrap resample the data
            bsample <- sample(data, n, replace = TRUE)

            # Recurse to the next depth, passing the reduced slice of levels
            result <- recurse(bsample, lvls[-1], α, par_mean)

            # Store the mean and standard error
            boot$μ[i] <- result$μ
            boot$se[i] <- result$std_err

            # Store the studentized t-statistic if requested
            if (studentized) boot$t_stat[i] <- (boot$μ[i] - par_mean) / boot$se[i]

            if (debug) cat("Iteration: ", i, "\n", sep = "")

        }

        # This only happens once at the top level
        if (len == lvls_len) return(list(μ = mean(boot$μ), lvls = lvls, α = α, par_mean = par_mean)) 
        # This happens at every other level
        else return(list(μ = mean(boot$μ), std_err = sd(boot$μ)))

    }

    return(recurse(data, lvls, α, par_mean))

}

for (name in names(ddat)) {

    if (name == "target") next
    col <- ddat[[name]]
    boot <- bootstrap(ddat$age, c(100, 100, 5), studentized = TRUE)
    mle <- boot$μ
    ci <- conf_int(col, mle)

    cat(

        "\n\nColumn:", name, "\n",
        "Bootstrap MLE: ", mle, "\n",
        "Normal CI: [", ci$nm[1], ", ", ci$nm[2], "]\n",
        "Likelihood CI: [", ci$ll[1], ", ", ci$ll[2], "]\n",
        sep = ""

    )
}
@
